Type theory in type theory with single substitutions

### 2024-09-16 types_2024_postproc/p.tex
stuff

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In the first order single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,i \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from a single substitution model.
TODO: universe levels are a problem (maybe not if we use iterated
function space instead of $\Sigma$); the resulting CwF is not the same as
the free CwF over the SSC. Bootstrapping a CwF.
CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
SSC -> CwF -> SSC = we get a stronger thing than what we started with.

@@

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument. Inrinsically typed terms are the same as derivation trees,
we illustrate this by deriving this term as follows.
\[
\infer[(*)]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
this is why we had to lift $\El\,\q$. Another subtle thing is
happening when deriving the third implicit argument of $\q$. We have
to coerce $\q$ along the equality $\U[]$.
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.

TODO: beta equation for id.

@@

@@

@@

@@

QIIT is nice because syntax and semantics are given together. The
notion of model/semantics is traditionally given separately, and work
has to be done to show that the above specified syntax generates a
model. For example, this was the goal of the initiality project for
homotopy type theory \cite{brunerie}. Maybe list work on type theory
using QIITs, see TYPES mailing list.

>>>>

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In the first order single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from a single substitution model.
TODO: universe levels are a problem (maybe not if we use iterated
function space instead of $\Sigma$); the resulting CwF is not the same as
the free CwF over the SSC. Bootstrapping a CwF.
CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
SSC -> CwF -> SSC = we get a stronger thing than what we started with.

@@

\subsection{Examples}

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument, and these arguments themselves are terms written using
implicit arguments. Inrinsically typed terms are the same as
derivation trees, we illustrate this by deriving this term as follows.
\[
\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
this is why we had to lift $\El\,\q$. Another subtle thing is
happening when deriving the third implicit argument of $\q$. We have
to coerce $\q$ along the equality $\U[]$.
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.

TODO: beta equation for id, show what we can do with types in an open
context.

\subsection{Models}

TODO: To make sure that our definitions of models are correct, we show
that in Agda there is a model for both the parallel and single
substitution calculi.

@@

TODO: summarise this section with a theorem?

@@

We can emulate the parallel substitution calculus using function
space, U and Sigma types

@@

We showed a particular example for the SOGAT->single GAT translation,
we showed that the syntax is equivalent to the other one, we
conjecture that the method presented here generalises. For this
particular SOGAT, the single GAT syntax can be simplified, and in
fact, it is ``equivalent'' to the parallel theory.

@@

QIIT is nice because syntax and semantics are given together. The
notion of model/semantics is traditionally given separately, and work
has to be done to show that the above specified syntax generates a
model. For example, this was the goal of the initiality project for
homotopy type theory \cite{initiality-agda}. Maybe list work on type theory
using QIITs, see TYPES mailing list.
